---
title: "Ames Housing Price Regression — Model Selection & Diagnostics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(scipen = 999)
```

# 1. Import and Prepare Data

```{r}
library(AmesHousing)
library(dplyr)
library(ggplot2)
library(MASS)
library(knitr)

ames <- make_ames()

# Only ONE predictor transformation is used: log(1 + Garage_Area).
# (Plain log fails because no-garage houses have Garage_Area = 0.)
cleaned_data <- ames |>
  mutate(
    Age_ames = Year_Sold - Year_Built,
    Age_ames = ifelse(Age_ames < 0, 0, Age_ames),
    Garage_Area_log1p = log1p(Garage_Area)
  ) |>
  dplyr::select(
    Sale_Price, Gr_Liv_Area, Total_Bsmt_SF, Garage_Area, Garage_Area_log1p,
    Lot_Area, Full_Bath, Bedroom_AbvGr, Fireplaces, Age_ames,
    TotRms_AbvGrd, Overall_Qual, Neighborhood
  ) |>
  na.omit() |>
  mutate(
    Overall_Qual = factor(Overall_Qual),
    Neighborhood = factor(Neighborhood),
    row_id = row_number()
  )

top15_neighborhoods <- cleaned_data |>
  count(Neighborhood, sort = TRUE) |> slice_head(n = 15) |>
  pull(Neighborhood) |> as.character()
```

```{r helpers}
# Coefficient table with 95% CI
coef_table <- function(model, conf = 0.95) {
  sm <- summary(model)$coefficients
  tcrit <- qt(1 - (1 - conf) / 2, model$df.residual)
  data.frame(Term = rownames(sm), Estimate = round(sm[, 1], 4),
             Std_Error = round(sm[, 2], 4),
             CI_Lower = round(sm[, 1] - tcrit * sm[, 2], 4),
             CI_Upper = round(sm[, 1] + tcrit * sm[, 2], 4),
             p_value = signif(sm[, 4], 4), row.names = NULL)
}

# Per-term partial F-test
partial_F_table <- function(full_model, terms_vec) {
  do.call(rbind, lapply(terms_vec, function(t) {
    reduced <- update(full_model, as.formula(paste(". ~ . -", t)))
    a <- anova(reduced, full_model)
    data.frame(Term = t, F = round(a$F[2], 3), p_value = signif(a$`Pr(>F)`[2], 4))
  }))
}

# Selection metrics with AICc
model_metrics <- function(model, label) {
  n <- nobs(model); p <- length(coef(model)) - 1
  aic <- extractAIC(model)[2]; bic <- extractAIC(model, k = log(n))[2]
  aicc <- aic + (2 * (p + 2) * (p + 3)) / (n - p - 1)
  data.frame(Model = label, Predictors = p,
             Adj_R2 = round(summary(model)$adj.r.squared, 4),
             AIC = round(aic, 2), AICc = round(aicc, 2), BIC = round(bic, 2),
             row.names = NULL)
}

# K-fold cross-validation MSE
cv_mse <- function(formula, data, response, K = 10, seed = 302) {
  set.seed(seed); n <- nrow(data); folds <- sample(rep(1:K, length.out = n)); err <- numeric(K)
  for (k in 1:K) {
    tr <- data[folds != k, ]; va <- data[folds == k, ]
    m <- lm(formula, data = tr); keep <- rep(TRUE, nrow(va))
    for (fac in names(m$xlevels)) keep <- keep & (as.character(va[[fac]]) %in% m$xlevels[[fac]])
    va <- va[keep, ]; if (nrow(va) == 0) { err[k] <- NA; next }
    err[k] <- mean((va[[response]] - predict(m, newdata = va))^2)
  }
  mean(err, na.rm = TRUE)
}

# Human-readable axis labels
pretty_label <- function(v) {
  lab <- c(
    Gr_Liv_Area       = "Above-Grade Living Area (sq ft)",
    Total_Bsmt_SF     = "Total Basement Area (sq ft)",
    Garage_Area       = "Garage Area (sq ft)",
    Garage_Area_log1p = "log(1 + Garage Area)",
    Lot_Area          = "Lot Area (sq ft)",
    Full_Bath         = "Full Bathrooms",
    Bedroom_AbvGr     = "Bedrooms Above Grade",
    Fireplaces        = "Number of Fireplaces",
    Age_ames          = "Age of House (years)",
    TotRms_AbvGrd     = "Total Rooms Above Grade",
    Overall_Qual      = "Overall Quality",
    Neighborhood      = "Neighborhood"
  )
  out <- unname(lab[v]); ifelse(is.na(out), v, out)
}

# Boxplot of residuals vs a categorical predictor, with x labels tilted 45 degrees.
tilted_boxplot <- function(resid, group, title, cex.axis = 0.75, bottom = 8) {
  group <- droplevels(as.factor(group))
  op <- par(mar = c(bottom, 4, 2.5, 1), mgp = c(2.5, 0.7, 0))
  bp <- boxplot(resid ~ group, xaxt = "n", xlab = "", ylab = "Residuals",
                main = title, col = "grey90", outcex = 0.6)
  abline(h = 0, col = "red")
  axis(1, at = seq_along(bp$names), labels = FALSE)           # ticks only
  text(x = seq_along(bp$names),
       y = par("usr")[3] - (par("usr")[4] - par("usr")[3]) * 0.07,
       labels = bp$names, srt = 45, adj = 1, xpd = TRUE, cex = cex.axis)
  par(op)
}

# Standard model-check panel for one model: Response vs Fitted, Residuals vs Fitted,
# Normal Q-Q, and a Fitted-value histogram. Works for any lm (raw or transformed
# response). Response vs Fitted: points should lie on the y = x line (blue).
check_plots <- function(model, label) {
  y    <- model.response(model.frame(model))
  yhat <- fitted(model)
  r    <- resid(model)
  op <- par(mfrow = c(2, 2), mar = c(3.4, 3.4, 2.4, 0.8), mgp = c(2, 0.6, 0))
  plot(yhat, y, main = paste(label, "Response vs Fitted"),
       xlab = "Fitted values", ylab = "Response"); abline(0, 1, col = "blue")
  plot(yhat, r, main = paste(label, "Residuals vs Fitted"),
       xlab = "Fitted values", ylab = "Residuals"); abline(h = 0, col = "red")
  qqnorm(r, main = paste(label, "Normal Q-Q")); qqline(r, col = "red")
  hist(yhat, main = paste(label, "Fitted Histogram"),
       xlab = "Fitted values", col = "grey85", border = "white")
  par(op)
}

# Residuals vs EVERY predictor in a model: numeric -> scatter grid, categorical ->
# tilted boxplot (Neighborhood restricted to its 15 most common levels).
residual_vs_predictors <- function(model) {
  mf <- model.frame(model)
  r  <- resid(model)
  preds <- attr(terms(model), "term.labels")
  num_preds <- preds[vapply(preds, function(t) is.numeric(mf[[t]]), logical(1))]
  cat_preds <- preds[vapply(preds, function(t) is.factor(mf[[t]]) || is.character(mf[[t]]), logical(1))]

  if (length(num_preds) > 0) {
    op <- par(mfrow = c(ceiling(length(num_preds) / 3), 3),
              mar = c(3.4, 3.4, 2.2, 0.8), mgp = c(2, 0.6, 0))
    for (x in num_preds) {
      plot(mf[[x]], r, main = paste("Residuals vs", pretty_label(x)),
           xlab = pretty_label(x), ylab = "Residuals"); abline(h = 0, col = "red")
    }
    par(op)
  }
  for (x in cat_preds) {
    g <- mf[[x]]
    if (x == "Neighborhood") {
      keep <- as.character(g) %in% top15_neighborhoods
      tilted_boxplot(r[keep], factor(g[keep], levels = top15_neighborhoods),
                     paste("Residuals vs", pretty_label(x), "(top 15)"),
                     cex.axis = 0.7, bottom = 9)
    } else {
      tilted_boxplot(r, g, paste("Residuals vs", pretty_label(x)),
                     cex.axis = 0.8, bottom = 7)
    }
  }
}
```

# 2. Step 1 — Initial (Preliminary) Model

```{r}
initial_model <- lm(
  Sale_Price ~ Gr_Liv_Area + Total_Bsmt_SF + Garage_Area +
    Age_ames + Overall_Qual + Neighborhood,
  data = cleaned_data
)

kable(coef_table(initial_model),
      caption = "Initial model: estimates, 95% CIs and p-values")
kable(partial_F_table(initial_model,
        c("Gr_Liv_Area", "Total_Bsmt_SF", "Garage_Area",
          "Age_ames", "Overall_Qual", "Neighborhood")),
      caption = "Initial model partial F-tests")
```

```{r, fig.width=8, fig.height=6}
check_plots(initial_model, "Initial Model")
```

```{r, fig.width=10, fig.height=7}
residual_vs_predictors(initial_model)
```

# 3. Step 2a — Model 1: Expanded Raw Model

`First_Flr_SF` / `Second_Flr_SF` are excluded (collinear with `Gr_Liv_Area`).

```{r}
model1 <- lm(
  Sale_Price ~ Gr_Liv_Area + Total_Bsmt_SF + Garage_Area + Lot_Area +
    Full_Bath + Bedroom_AbvGr + Fireplaces + Age_ames + TotRms_AbvGrd +
    Overall_Qual + Neighborhood,
  data = cleaned_data
)
kable(coef_table(model1), caption = "Model 1 (raw response): coefficient table")
```

```{r, fig.width=8, fig.height=6}
check_plots(model1, "Model 1")
```

```{r, fig.width=10, fig.height=7}
residual_vs_predictors(model1)
```

# 4. Step 2b — Box-Cox (Response Transformation)

```{r, fig.width=6, fig.height=4}
boxcox_result <- boxcox(model1, lambda = seq(-2, 2, 0.05))
lambda <- boxcox_result$x[which.max(boxcox_result$y)]
ci_mask <- boxcox_result$y > (max(boxcox_result$y) - qchisq(0.95, 1) / 2)
lambda_ci <- range(boxcox_result$x[ci_mask])
```

```{r}
kable(data.frame(
  Quantity = c("Estimated lambda", "95% CI lower", "95% CI upper"),
  Value = round(c(lambda, lambda_ci[1], lambda_ci[2]), 3)),
  caption = "Box-Cox: ML estimate of lambda and 95% CI")
```

```{r}
lambda_use <- lambda
geom_mean <- exp(mean(log(cleaned_data$Sale_Price)))   # = prod(y)^(1/n), overflow-safe

if (abs(lambda_use) < 1e-8) {
  bc_forward <- function(y) geom_mean * log(y)
  bc_inverse <- function(z) exp(z / geom_mean)
} else {
  bc_forward <- function(y) geom_mean^(1 - lambda_use) * (y^lambda_use - 1) / lambda_use
  bc_inverse <- function(z) ((z * lambda_use) / geom_mean^(1 - lambda_use) + 1)^(1 / lambda_use)
}
cleaned_data <- cleaned_data |> mutate(Sale_Price_boxcox = bc_forward(Sale_Price))
```

# 5. Step 2c — Model 2: Transformed Response, Raw Predictors

```{r}
model2 <- lm(
  Sale_Price_boxcox ~ Gr_Liv_Area + Total_Bsmt_SF + Garage_Area + Lot_Area +
    Full_Bath + Bedroom_AbvGr + Fireplaces + Age_ames + TotRms_AbvGrd +
    Overall_Qual + Neighborhood,
  data = cleaned_data
)
model2_numeric <- c("Gr_Liv_Area", "Total_Bsmt_SF", "Garage_Area", "Lot_Area",
                    "Full_Bath", "Bedroom_AbvGr", "Fireplaces", "Age_ames", "TotRms_AbvGrd")
kable(coef_table(model2), caption = "Model 2 (transformed response): coefficient table")
```

```{r, fig.width=8, fig.height=6}
check_plots(model2, "Model 2")
```

```{r, fig.width=10, fig.height=7}
residual_vs_predictors(model2)
```

# 6. Step 2d — Model 3: Single Predictor Transformation (log Garage)

The only predictor transformation kept is `log(1 + Garage_Area)`, motivated by the
curvature/right skew in the Garage_Area residual plot. All other predictors stay on
their raw scale; the response stays Box-Cox transformed.

```{r}
model3 <- lm(
  Sale_Price_boxcox ~ Gr_Liv_Area + Total_Bsmt_SF + Garage_Area_log1p + Lot_Area +
    Full_Bath + Bedroom_AbvGr + Fireplaces + Age_ames + TotRms_AbvGrd +
    Overall_Qual + Neighborhood,
  data = cleaned_data
)
model3_numeric <- c("Gr_Liv_Area", "Total_Bsmt_SF", "Garage_Area_log1p", "Lot_Area",
                    "Full_Bath", "Bedroom_AbvGr", "Fireplaces", "Age_ames", "TotRms_AbvGrd")

kable(coef_table(model3), caption = "Model 3 (single transform): coefficient table")
```

```{r}
vif_model3 <- c()
for (x in model3_numeric) {
  other_x <- setdiff(model3_numeric, x)
  vif_lm <- lm(as.formula(paste(x, "~", paste(other_x, collapse = " + "))), data = cleaned_data)
  vif_model3[x] <- 1 / (1 - summary(vif_lm)$r.squared)
}
kable(data.frame(Predictor = names(vif_model3), VIF = round(as.numeric(vif_model3), 3)) |>
        arrange(desc(VIF)),
      caption = "Model 3 VIF check (cutoff: VIF > 5)")
```

```{r, fig.width=8, fig.height=6}
check_plots(model3, "Model 3")
```

```{r, fig.width=10, fig.height=7}
residual_vs_predictors(model3)
```

# 7. Step 3 — Remove Outliers, Bad Leverage, and Influential Points

Diagnostics are computed on Model 3. Cutoffs used:

* **High leverage**: $h_{ii} > 2(p+1)/n$.
* **Outlier**: $|r_i| > 4$ (standardized residual).
* **Bad leverage**: high leverage **and** outlying residual.
* **Influential**: Cook's distance $D_i > 4/n$
  $\left(D_i=\frac{r_i^2}{p+1}\cdot\frac{h_{ii}}{1-h_{ii}}\right)$.

We remove the union of **bad leverage + outliers + influential** observations.

```{r}
candidate_model <- model3
n <- nobs(candidate_model)
p_plus_1 <- length(coef(candidate_model))

leverage_cutoff <- 2 * p_plus_1 / n
outlier_cutoff  <- 4          # |standardized residual| > 4
cooks_cutoff    <- 4 / n      # influential: Cook's D > 4/n

obs_diag <- cleaned_data |>
  mutate(
    std_resid = rstandard(candidate_model),
    leverage  = hatvalues(candidate_model),
    cooks_d   = cooks.distance(candidate_model)
  ) |>
  mutate(
    high_leverage = !is.na(leverage)  & leverage > leverage_cutoff,
    outlier       = !is.na(std_resid) & abs(std_resid) > outlier_cutoff,
    bad_leverage  = high_leverage & outlier,
    influential   = !is.na(cooks_d)  & cooks_d > cooks_cutoff,
    removal_candidate = bad_leverage | outlier | influential
  )

kable(data.frame(
  Diagnostic = c("High leverage (h_ii > 2(p + 1)/n)", "Outlier (|std resid| > 4)",
                 "Bad leverage (high lev & outlier)", "Influential (Cook's D > 4/n)",
                 "Removed (union of bad leverage + outliers + influential)"),
  Cutoff = round(c(leverage_cutoff, outlier_cutoff, leverage_cutoff, cooks_cutoff, NA), 4),
  Count  = c(sum(obs_diag$high_leverage), sum(obs_diag$outlier),
             sum(obs_diag$bad_leverage), sum(obs_diag$influential),
             sum(obs_diag$removal_candidate))),
  caption = "Problematic-observation counts (Model 3)")
```

```{r}
removal_ids <- obs_diag |> filter(removal_candidate) |> pull(row_id)
data_clean  <- cleaned_data |> filter(!(row_id %in% removal_ids))
data_clean  <- droplevels(data_clean)

kable(data.frame(
  Quantity = c("Removed observations", "Remaining observations"),
  Count    = c(length(removal_ids), nrow(data_clean))),
  caption = "Effect of removing flagged observations")
```

# 8. Step 4 — Backward Elimination (step by step)

All selection is by **backward elimination using AIC**. Starting from the
full transformed model, at each step we drop the single term whose removal lowers AIC
the most, stopping when no removal improves AIC. Each intermediate model is recorded
and checked.

```{r}
full_model_clean <- lm(
  Sale_Price_boxcox ~ Gr_Liv_Area + Total_Bsmt_SF + Garage_Area_log1p + Lot_Area +
    Full_Bath + Bedroom_AbvGr + Fireplaces + Age_ames + TotRms_AbvGrd +
    Overall_Qual + Neighborhood,
  data = data_clean
)

backward_models <- list(full_model_clean)
backward_log <- data.frame()
current <- full_model_clean
repeat {
  terms_now <- attr(terms(current), "term.labels")
  if (length(terms_now) <= 1) break
  base_aic <- extractAIC(current)[2]
  drop_aic <- sapply(terms_now, function(t)
    extractAIC(update(current, as.formula(paste(". ~ . -", t))))[2])
  if (min(drop_aic) >= base_aic) break          # no removal improves AIC -> stop
  drop_term <- names(which.min(drop_aic))
  current <- update(current, as.formula(paste(". ~ . -", drop_term)))
  backward_models[[length(backward_models) + 1]] <- current
  backward_log <- rbind(backward_log, data.frame(
    Step = nrow(backward_log) + 1, Dropped_term = drop_term,
    AIC_after = round(min(drop_aic), 2),
    Adj_R2 = round(summary(current)$adj.r.squared, 4)))
}
final_model <- current
final_terms <- attr(terms(final_model), "term.labels")
```

```{r}
if (nrow(backward_log) == 0) {
  kable(data.frame(Note = "No term removal lowered AIC; the full model is final."),
        caption = "Backward elimination path")
} else {
  kable(backward_log, caption = "Backward elimination path (term dropped at each AIC step)")
}

kable(data.frame(Final_terms = final_terms), caption = "Final model retained terms")

# Metrics across every backward step
comp <- do.call(rbind, lapply(seq_along(backward_models), function(i) {
  model_metrics(backward_models[[i]], if (i == 1) "Full model" else paste("After step", i - 1))
}))
kable(comp, caption = "Selection criteria across backward elimination steps")
```

# 9. Model Checking for Each Backward Step

Each model produced during backward elimination receives the same diagnostic checks.

```{r, fig.width=10, fig.height=7, results='asis'}
# Each backward model gets the 4-panel check AND residuals vs every predictor.
# (If the report gets too long, restrict to the full model and final model with:
#  for (i in c(1, length(backward_models))) { ... } )
for (i in seq_along(backward_models)) {
  lab <- if (i == 1) "Full model" else paste("Backward Step", i - 1)
  check_plots(backward_models[[i]], lab)
  residual_vs_predictors(backward_models[[i]])
}
```

# 10. Final Model: VIF, Partial F, and Assumption Checks

```{r}
all_numeric <- c("Gr_Liv_Area", "Total_Bsmt_SF", "Garage_Area_log1p", "Lot_Area",
                 "Full_Bath", "Bedroom_AbvGr", "Fireplaces", "Age_ames", "TotRms_AbvGrd")
final_numeric <- intersect(all_numeric, final_terms)
final_categorical <- intersect(c("Overall_Qual", "Neighborhood"), final_terms)

if (length(final_numeric) > 1) {
  vif_final <- c()
  for (x in final_numeric) {
    other_x <- setdiff(final_numeric, x)
    vif_lm <- lm(as.formula(paste(x, "~", paste(other_x, collapse = " + "))), data = data_clean)
    vif_final[x] <- 1 / (1 - summary(vif_lm)$r.squared)
  }
  kable(data.frame(Predictor = names(vif_final), VIF = round(as.numeric(vif_final), 3)) |>
          arrange(desc(VIF)), caption = "Final model VIF values")
}

kable(partial_F_table(final_model, final_terms),
      caption = "Final model partial F-tests")
```

```{r, fig.width=8, fig.height=6}
check_plots(final_model, "Final Model")
```

```{r, fig.width=10, fig.height=7}
residual_vs_predictors(final_model)
```


# 11. Present Final Model (vs Initial Model)

The final model is compared with the initial (proposal) model fit on the same cleaned,
transformed data. This confirms whether keeping only one predictor transformation still
improves on the proposal model in AIC/BIC.

```{r}
initial_bc <- lm(
  Sale_Price_boxcox ~ Gr_Liv_Area + Total_Bsmt_SF + Garage_Area +
    Age_ames + Overall_Qual + Neighborhood,
  data = data_clean
)

kable(rbind(
  model_metrics(initial_bc,  "Initial model"),
  model_metrics(final_model, "Final model")),
  caption = "Initial vs final model (lower AIC/AICc/BIC, higher Adj-R^2 is better)")

kable(coef_table(final_model),
      caption = "Final model: estimates, 95% CIs and p-values")
```

# 12. Step 5 — Split Data (after model selection)

```{r}
set.seed(302)
rows  <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
train <- data_clean[rows, ]
test  <- data_clean[-rows, ]
```

# 13. Step 6 — Train, Validate, and Compare to Initial Model

```{r}
fit_train_validate <- function(form, tr, te, response) {
  m <- lm(form, data = tr); keep <- rep(TRUE, nrow(te))
  for (fac in names(m$xlevels)) keep <- keep & (as.character(te[[fac]]) %in% m$xlevels[[fac]])
  te <- te[keep, ]; pred <- predict(m, newdata = te)
  list(model = m, n_used = nrow(te),
       rmse_t = sqrt(mean((te[[response]] - pred)^2)),
       rmse_price = sqrt(mean((te$Sale_Price - bc_inverse(pred))^2)))
}

val_final   <- fit_train_validate(formula(final_model), train, test, "Sale_Price_boxcox")
val_initial <- fit_train_validate(formula(initial_bc),  train, test, "Sale_Price_boxcox")
cv_final   <- cv_mse(formula(final_model), data_clean, "Sale_Price_boxcox", K = 10)
cv_initial <- cv_mse(formula(initial_bc),  data_clean, "Sale_Price_boxcox", K = 10)

kable(data.frame(
  Model = c("Initial model", "Final model"),
  Test_n_used = c(val_initial$n_used, val_final$n_used),
  Test_RMSE_transformed = round(c(val_initial$rmse_t, val_final$rmse_t), 4),
  CV10_RMSE_transformed = round(c(sqrt(cv_initial), sqrt(cv_final)), 4),
  Test_RMSE_price = round(c(val_initial$rmse_price, val_final$rmse_price), 2)),
  caption = "Validation: final vs initial model (lower RMSE is better)")

kable(coef_table(val_final$model),
      caption = "Final model coefficients (beta-hat) estimated on the training set")
```
